Derogations, Democratic Backsliding, and International Human Rights During the COVID-19 Pandemic

### 2024-11-25 manuscript/manuscript.qmd
Address R&R issue 6

---
title: "Derogations, Democratic Backsliding, and International Human Rights During the COVID-19 Pandemic"
short-title: "Derogations and Backsliding"
author:
- name: Suparna Chaudhry
  affiliation: Lewis and Clark College
  orcid: 0000-0003-3506-7590
  url: https://www.suparnachaudhry.com/
  email: schaudhry@lclark.edu
- name: Audrey L. Comstock
  affiliation: Arizona State University
  orcid: 0000-0003-1772-5350
  url: https://audreylcomstock.weebly.com/
  email: audrey.comstock@asu.edu
- name: Andrew Heiss
  affiliations:
    - id: gsu
      name: Georgia State University
      department: Andrew Young School of Policy Studies, Department of Public Management and Policy
      address: "55 Park Place NE, #464"
      city: Atlanta
      region: GA
      country: USA
      postal-code: 30303
  orcid: 0000-0002-3948-3914
  url: https://www.andrewheiss.com
  email: aheiss@gsu.edu
  attributes:
    corresponding: true
correspondence-prefix: "Correspondence concerning this article should be addressed to"
abstract: |
  Did states misuse international legal emergency provisions during the COVID-19 pandemic to justify human rights abuse or did they follow international human rights law? Many governments restricted citizens' freedom of movement, association, and assembly during the crisis, raising questions about states' commitments to international human rights law. Some states used derogations to communicate temporary suspension of international legal provisions in a proportional and non-discriminatory manner, while others did not. We explore the dynamics of democratic backsliding and derogation use during the pandemic. We find that backsliding states were more likely to issue human rights treaty derogations. These derogations had mitigating effects once issued. Backsliding states that issued derogations were more likely to communicate restrictions and were less likely to issue abusive and discriminatory policy during the pandemic. Derogations helped temper abuse in states not experiencing backsliding. However, derogations did not always protect against abuse and media transparency in backsliding states. These results lend support to the use of flexibility mechanisms in international law and find that most states did not use emergency derogations to heighten human rights violations. The study contributes to the understanding of how international legal measures may help mitigate elements of democratic backsliding during times of crisis.
date: 2024-11-24
code-repo: "Access the code, data, and analysis at <https://github.com/andrewheiss/lemon-lucifer>"
keywords:
  - human rights
  - international law
  - pandemic
  - derogation
  - democratic backsliding
bibliography: references.json
csl: csl/apa.csl
link-citations: true
---

@@

@@

Global democratic backsliding was already in motion when the pandemic spread in March 2020 [@LuhrmannLindberg:2020; @MaerzLuhrmannHellmeier:2020]. As of 2019, the share of countries experiencing democratic erosion more than doubled in the past decade compared to the decade before [@IDEA:2019], with only 8% of the world's population living in countries becoming more democratic [@LuhrmannLindberg:2020]. Democratic backsliding is a particular subset of such erosion involving the intentional weakening of checks and balances, and the curtailment and rollbacks of civil liberties, political rights, personal freedoms, and a broad swath of human rights. @fig-panback-map demonstrates that North America and Western Europe generally had low risk of democratic backsliding during the pandemic. However, this does not indicate that there was no backsliding occurring. Researchers point to election manipulation and executive overreach in the United States as examples of democratic erosion [@Williamson:2023] and the European Union has responded via sanctions to curb democratic backsliding of member states [@BlaubergerSedelmeier:2024]. The higher risk areas included parts of South America, across Africa, India, and some parts of Eastern Europe and Southeast Asia. Overall, pandemic-era democratic backsliding was a global phenomenon. @fig-derogation-map depicts states that issued pandemic-related derogations to the ICCPR. Most states do not issue derogations even during times of emergency, so the determinants of when they are issued are important to understand. Geographically, derogations were issued across different regions including South America, Africa, Eastern Europe, and some parts of Asia. They were not generally issued across North America or Western Europe.

>>>>

---
title: "Derogations, Democratic Backsliding, and International Human Rights During the COVID-19 Pandemic"
short-title: "Derogations and Backsliding"
author:
- name: Suparna Chaudhry
  affiliation: Lewis and Clark College
  orcid: 0000-0003-3506-7590
  url: https://www.suparnachaudhry.com/
  email: schaudhry@lclark.edu
- name: Audrey L. Comstock
  affiliation: Arizona State University
  orcid: 0000-0003-1772-5350
  url: https://audreylcomstock.weebly.com/
  email: audrey.comstock@asu.edu
- name: Andrew Heiss
  affiliations:
    - id: gsu
      name: Georgia State University
      department: Andrew Young School of Policy Studies, Department of Public Management and Policy
      address: "55 Park Place NE, #464"
      city: Atlanta
      region: GA
      country: USA
      postal-code: 30303
  orcid: 0000-0002-3948-3914
  url: https://www.andrewheiss.com
  email: aheiss@gsu.edu
  attributes:
    corresponding: true
correspondence-prefix: "Correspondence concerning this article should be addressed to"
abstract: |
  Did states misuse international legal emergency provisions during the COVID-19 pandemic to justify human rights abuse or did they follow international human rights law? Many governments restricted citizens' freedom of movement, association, and assembly during the crisis, raising questions about states' commitments to international human rights law. Some states used derogations to communicate temporary suspension of international legal provisions in a proportional and non-discriminatory manner, while others did not. We explore the dynamics of democratic backsliding and derogation use during the pandemic. We find that backsliding states were more likely to issue human rights treaty derogations. These derogations had mitigating effects once issued. Backsliding states that issued derogations were more likely to communicate restrictions and were less likely to issue abusive and discriminatory policy during the pandemic. Derogations helped temper abuse in states not experiencing backsliding. However, derogations did not always protect against abuse and media transparency in backsliding states. These results lend support to the use of flexibility mechanisms in international law and find that most states did not use emergency derogations to heighten human rights violations. The study contributes to the understanding of how international legal measures may help mitigate elements of democratic backsliding during times of crisis.
date: 2024-11-25
code-repo: "Access the code, data, and analysis at <https://github.com/andrewheiss/lemon-lucifer>"
keywords:
  - human rights
  - international law
  - pandemic
  - derogation
  - democratic backsliding
bibliography: references.json
csl: csl/apa.csl
link-citations: true
---

@@

::: {.content-visible when-format="html"}
# Introduction {.hide}
:::

@@

Global democratic backsliding was already in motion when the pandemic spread in March 2020 [@LuhrmannLindberg:2020; @MaerzLuhrmannHellmeier:2020]. As of 2019, the share of countries experiencing democratic erosion more than doubled in the past decade compared to the decade before [@IDEA:2019], with only 8% of the world's population living in countries becoming more democratic [@LuhrmannLindberg:2020]. Democratic backsliding is a particular subset of such erosion involving the intentional weakening of checks and balances, and the curtailment and rollbacks of civil liberties, political rights, personal freedoms, and a broad swath of human rights, in any state, regardless of regime type.[^backsliding] @fig-panback-map demonstrates that North America and Western Europe generally had low risk of democratic backsliding during the pandemic. However, this does not indicate that there was no backsliding occurring. Researchers point to election manipulation and executive overreach in the United States as examples of democratic erosion [@Williamson:2023] and the European Union has responded via sanctions to curb democratic backsliding of member states [@BlaubergerSedelmeier:2024]. The higher risk areas included parts of South America, across Africa, India, and some parts of Eastern Europe and Southeast Asia. Overall, pandemic-era democratic backsliding was a global phenomenon. @fig-derogation-map depicts states that issued pandemic-related derogations to the ICCPR. Most states do not issue derogations even during times of emergency, so the determinants of when they are issued are important to understand. Geographically, derogations were issued across different regions including South America, Africa, Eastern Europe, and some parts of Asia. They were not generally issued across North America or Western Europe.

[^backsliding]: The definition of democratic backsliding is contested. Some scholars claim that backsliding can only occur in existing democracies, while others argue that it can happen in any state, regardless of regime type [@LittleMeng:2024]. Following other research based on V-Dem's PanDem data, we take a maximalist definition of backsliding from @WaldnerLust:2018, who argue that "backsliding entails a deterioration of qualities associated with democratic governance, within any regime. In democratic regimes, it is a decline in the quality of democracy; in autocracies, it is a decline in democratic qualities of governance" [-@WaldnerLust:2018, 95].
